Modal Logics in the Coq Proof Assistant

### 2014-12-09 Papers/InProgress/ModalLogicCoq/ModalCoq.tex
Write section 2 (The Embedding in Coq)

\section{Modal Logic in Coq}
\label{sec:modal}

A crucial aspect of modal logics \cite{ModalLogic} is that the so-called \emph{necessitation rule} allows $\Box A$ to be derived if $A$ is a theorem, but $A \imp \Box A$ is not necessarily a theorem. Naive attempts to define the modal operators $\Box$ and $\Diamond$ may easily be unsound in this respect. To avoid this issue, the \emph{possible world semantics} of modal logics can be explicitly embedded into higher-order logics \cite{J23,B9}.

The embedding is related to labeling techniques
\cite{Labels}. However, the expressiveness of higher-order logic is
exploited here to encode the labels within the logic language itself. To this aim, a type for worlds must be declared and modal propositions should be not of type \texttt{Prop} but of a lifted type \texttt{o} that depends on possible worlds:

@@

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Definition mequal (x y: u) := \fun w: i => x = y.
\Notation "x m= y" := (mequal x y) (at level 99, right associativity).

@@

\noindent
Likewise, modal quantifiers are lifted versions of the usual quantifiers.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Definition A {t: Type}(p: t -> o) := \fun w => \fforall x, p x w.
\Notation "'mforall'  x , p" := (A (fun x => p))
  (at level 200, x ident, right associativity) : type_scope.
\Notation "'mforall' x : t , p" := (A (\fun x:t => p))
  (at level 200, x ident, right associativity, 
    format "'[' 'mforall' '/ '  x  :  t , '/ '  p ']'")
  : type_scope.

\Definition E {t: Type}(p: t -> o) := \fun w => exists x, p x w.
\Notation "'mexists' x , p" := (E (\fun x => p))
  (at level 200, x ident, right associativity) : type_scope.
\Notation "'mexists' x : t , p" := (E (\fun x:t => p))
  (at level 200, x ident, right associativity, 
    format "'[' 'mexists' '/ '  x  :  t , '/ '  p ']'")
  : type_scope.
\end{Verbatim}

\noindent
The modal operators $\Diamond$ (\emph{possibly}) and $\Box$ (\emph{necessarily}) are defined accordingly to their meanings in the possible world semantics.

>>>>

\section{The Embedding of Modal Logics in Coq}
\label{sec:modal}

A crucial aspect of modal logics \cite{ModalLogic} is that the so-
called \emph{necessitation rule} allows $\Box A$ to be derived if $A$
is a theorem, but $A \imp \Box A$ is not necessarily a theorem. Naive
attempts to define the modal operators $\Box$ and $\Diamond$ may
easily be unsound in this respect. To avoid this issue, the
\emph{possible world semantics} of modal logics can be explicitly
embedded into higher-order logics \cite{J23,B9}.

The embedding is related to labeling techniques \cite{Labels}.
However, the expressiveness of higher-order logic is exploited here to
encode the labels within the logical language. To this aim, a type for
worlds must be declared and modal propositions should be not of type
\texttt{Prop} but of a lifted type \texttt{o} that depends on possible
worlds:

@@

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Definition mequal (x y: u)(w: i) := x = y.
\Notation "x m= y" := (mequal x y) (at level 99, right associativity).

@@

\noindent 
Likewise, modal quantifiers are lifted versions of the usual
quantifiers. \Coq's type system with dependent types is particularly
helpful here. The modal quantifiers \texttt{A} and \texttt{E} are
defined as depending on a type \texttt{t}. Therefore, they can
quantify over variables of any type. Moreover, the curly brackets
indicate that \texttt{t} is an implicit argument that can be inferred
by \Coq's type inference mechanism. This allows notations\footnote{The
keyword \texttt{\fun} indicates a lambda abstraction: \texttt{\fun \ x
=> p} (or \texttt{\fun \ x:t => p}) denotes the function 
$\lambda x:t.p$, which takes an argument $x$ (of type $t$) and returns $p$.}
(i.e. \texttt{mforall} and \texttt{mexists}) that mimic the notations
for \Coq's usual quantifiers (i.e. \texttt{forall} and
\texttt{exists}).

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Definition A \{t: Type\}(p: t -> o)(w: i) := \fforall x, p x w.
\Notation "'mforall'  x , p" := (A (\fun x => p))
  (at level 200, x ident, right associativity) : type_scope.
\Notation "'mforall' x : t , p" := (A (\fun x:t => p))
  (at level 200, x ident, right associativity, 
    format "'[' 'mforall' '/ '  x  :  t , '/ '  p ']'")
  : type_scope.

\Definition E \{t: Type\}(p: t -> o)(w: i) := exists x, p x w.
\Notation "'mexists' x , p" := (E (\fun x => p))
  (at level 200, x ident, right associativity) : type_scope.
\Notation "'mexists' x : t , p" := (E (\fun x:t => p))
  (at level 200, x ident, right associativity, 
    format "'[' 'mexists' '/ '  x  :  t , '/ '  p ']'")
  : type_scope.
\end{Verbatim}

\noindent 
The modal operators $\Diamond$ (\emph{possibly}) and $\Box$
(\emph{necessarily}) are defined accordingly to their meanings in the
possible world semantics. $\nec p$ holds at a world $w$ iff $p$ holds
in every world $w_1$ reachable from $w$. $\pos p$ holds at world $w$
iff $p$ holds in some world $w_1$ reachable from $w$.
